---
title: "Tutorial for the Emil package"
author: Detlef Groth, University of Potsdam
date: 2023-02-12
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

## Introduction

The Emil package is an easy to use object oriented programming approach for R
in the spirit of Pythons object oriented syntax and as well similar to the
[R6](https://adv-r.hadley.nz/r6.html) system in R. It provides a single small
R file which can be very easily embedded into your own code with minimal
effort.

As an example of how to use the package we will create an Animal class/object
first.


```{r}
library(Emil)
Animal = Emil$new(age=0,type="generic")
Animal$getAge <- function (self) {
    return(self$age)
}
Animal$birthday <- function (self) {
    self$age=self$age+1
}
```

The Emil package does not distinguish between class and object, just to follow
standard convention we will uppercase class names and lower case names for objects. Let's create a first animal.

```{r}
lucky=Animal$new(name="lucky",type="sheep")
lucky$getAge()
lucky$birthday()
lucky$getAge()
```

At any time we can inspect our object using the standard `ls` and `str` functions of R:

```{r}
ls(lucky)
str(lucky)
```

## Inheritance

Like we derive objects we can as well create new templates for other objects. Let's create a cat class based on Animal.

```{r}
Cat = Animal$new(name="Miez")
ls(Cat)
```

We can then extend this class:

```{r}
Cat$meow <- function (self,x=3) {
    m=rep("meow",x)
    return(paste(self$name," says: ", paste(m, collapse=", "),"!",sep=""))
}
micky = Cat$new(name="Micky")
micky$meow(4)
```

## S3 methods

You can as well create standard S3 methods if you give your objects a new
class name. Let's create dogs with this approach.

```{r}
Dog = Animal$new(class="Dog",name="Generic")
class(Dog)
Dog$wuff <- function (self) {
    return("Wuff, wuff!")
}
fido = Dog$new(name="Fido")
class(fido)
fido$birthday()
fido$getAge()
fido$wuff()
```

Let's now create a S3 method print for Dogs:

```{r}
print.Dog <- function (x,...) {
    print(paste("I am a dog with name ",x$name,"!",sep=""))
}   
print(Dog)
print(fido)
```

You can as well attach the class name like this to existing objects or classes
at a later point if not given in the new call:

```{r}
class(micky)=c("Emil","Cat")
print.Cat <- function (x,...) {
    print(paste("I am",x$name, "the cat!"))
    x$meow()
}
print(micky)
```

## Constructor

Sometimes you like to execute some code at object creation, for instance
intialize some internal objects based on the given arguments. For this purpose
there is the argument list at object creation checked for the presence of a
function called `.init` (leading dot) which this function is then executed with the argument
self. Here an example:

```{r}
Graph = Emil$new(type="random",
    .init=function (self) {
        g=matrix(0,nrow=5,ncol=5)  
        rownames(g)=colnames(g)=LETTERS[1:5]
        if (self$type=="random") {
            g[upper.tri(g)]=rbinom(10,1,p=0.4)
        } else {
            g['A',c(2:5)]=1
        }
        self$graph=g
   })
g=Graph$new()   
g$graph
h=Graph$new(type="random")
h$graph
```
            
You could as well create the `.init` function after the new call like this:

```{r}
Graph = Emil$new(type="random")
Graph$.init <- function (self) {
     print("calling init")
     g=matrix(0,nrow=5,ncol=5)  
        rownames(g)=colnames(g)=LETTERS[1:5]
        if (self$type=="random") {
            g[upper.tri(g)]=rbinom(10,1,p=0.4)
        } else {
            g['A',c(2:5)]=1
        }
        self$graph=g

}
i = Graph$new()
ls(i)
i$graph
```

The latter approach as well calls the `.init` function automatically if new is executed.

## Accumulator example

Let's redo some parts of Hadley's
[Accumulator](https://adv-r.hadley.nz/r6.html) example using Emil:

```{r}
Accumulator <- Emil$new(sum=0)
Accumulator$add <- function (self,x=1) {
     self$sum <- self$sum + x 
     invisible(self) # for method chaining
}  
```

Ok now let's create an object:

```{r}
x <- Accumulator$new() 
x$add(4)
x$sum
```

Method chaining does work as well:

```{r}
x$add(10)$add(10)$sum
```

What is the sum?

```{r}
x$sum
```

Protecting internal variables against access from the outside does not exists
in Emil. You should prefix internal variables with a dot to indicate that they should not be modified from the outside.

```{r}
x$.intern = "hello"
ls(x)
```

As you you can see using the ls without the argument the variables starting
with a dot are not visible. But you can still modify and access the if you
like ... although you should not do this:

```{r}
x$.intern
x$.intern = "world"
x$.intern
```

## Documenting Emil objects/classes

Let's take the example above to demonstrate how you could document your R
class created with Emil.
 
Here the file `Accumulator-class.Rd`:

```
\docType{class}
\name{Accumulator-class}
\alias{Accumulator}
\alias{Accumulator-class}
\title{Example Accumulator class for Emil.}
\description{
An example class for the Emil OOP system.
}
\section{Fields}{
  \describe{
     \item{\code{sum}}{The current amount of the accumlator values}
  }
}
\section{Methods}{
\itemize{
   \item \code{add(x=1)}: Add the value of x to the current accumulated value in sum
}}
\examples{
 x = Accumulator$new()
 x$add(5)
 x$sum
}
```

The methods can be as well documented like this but the dollar should be
replaced then by an underline for one of the aliases, your filename should be
then as well have this underline.

Here the file `Accumulator_add.Rd`:

```
\name{Accumulator$add}
\alias{Accumulator_add}
\alias{Accumulator$add}
\title{Add the value of x to the current accumulated value in sum.}
\description{This is a method of the Accumulator class which adds the given value to the current object}  
\usage{`Accumulator$add(x=1)`}
\arguments{
    \item{x}{the value to be added to the current internal accumulator value sum, default: 1}
}
\value{object itself invisibly}    
\examples{
 y = Accumulator()
 y$add(5)
 y$sum
 y$add(3)$add(4)
 y$sum
}
```

To see the appropiate manual pages look at the files
`man/Accumulator-class.Rd` and `man/Accumulator_add.Rd`.


