---
title: "Tutorial for the Emil package"
author: Detlef Groth, University of Potsdam
date: 2023-02-12
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE,results="hide"}
library(knitr)
```

## Introduction

The Emil package is an easy to use object oriented programming approach for R
in the spirit of Pythons object oriented syntax and as well similar to the
[R6](https://adv-r.hadley.nz/r6.html) system in R. It provides a single small
R file which can be very easily embedded into your own code with minimal
effort.

As an example of how to use the package we will create an Animal class/object
first.


```{r}
library(Emil)
Animal = Emil$new(age=0,type="generic")
Animal$getAge <- function (self) {
    return(self$age)
}
Animal$birthday <- function (self) {
    self$age=self$age+1
}
```

The Emil package does not distinguish between class and object, just to follow
standard convention we will uppercase class names and lower case objects. Let's create a first animal.

```{r}
lucky=Animal$new(name="lucky",type="sheep")
lucky$getAge()
lucky$birthday()
lucky$getAge()
```

At any time we can inspect our object using the standard `ls` and `str` functions of R:

```{r}
ls(lucky)
str(lucky)
```

## Inheritance

Like we derive objects we can as well create new templates for other objects. Let's create a cat class based on Animal.

```{r}
Cat = Animal$new(name="Miez")
ls(Cat)
```

We can then extend this class:

```{r}
Cat$meow <- function (self,x=3) {
    m=rep("meow",x)
    return(paste(self$name," says: ", paste(m, collapse=", "),"!",sep=""))
}
micky = Cat$new(name="Micky")
micky$meow(4)
```

## S3 methods

You can as well create standard S3 methods if you give your objects a new
class name. Let's create dogs with this approach.

```{r}
Dog = Animal$new(class="Dog",name="Generic")
class(Dog)
Dog$wuff <- function (self) {
    return("Wuff, wuff!")
}
fido = Dog$new(name="Fido")
class(fido)
fido$birthday()
fido$getAge()
fido$wuff()
```

Let's now create a S3 method print for Dogs:

```{r}
print.Dog <- function (x,...) {
    print(paste("I am a dog with name ",x$name,"!",sep=""))
}   
print(Dog)
print(fido)
```

## Accumulator example

Let's redo some parts of Hadley's [Accumulator](https://adv-r.hadley.nz/r6.html) example using Emil:

```{r}
Accumulator <- Emil$new(sum=0)
Accumulator$add <- function (self,x=1) {
     self$sum <- self$sum + x 
     invisible(self) # for method chaining
}  
```

Ok now let's create an object:

```{r}
x <- Accumulator$new() 
x$add(4)
x$sum
```

Method chaining does work as well:

```{r}
x$add(10)$add(10)$sum
```

What is the sum?

```{r}
x$sum
```

Protecting internal variables against access from the outside does not exists
in Emil. You should prefix internal variables with a dot to indicate that they should not be modified from the outside.

```{r}
x$.intern = "hello"
ls(x)
```

As you you can see using the ls without the argument the variables starting
with a dot a re not visible. But you can still modify and access the if you
like ...

```{r}
x$.intern
x$.intern = "world"
x$.intern
```

## Documenting Emil objects/classes

Let's take the example above to demonstrate how you could document your R
class created with Emil.
 

```
\docType{class}
\name{Accumulator-class}
\alias{Accumulator}
\alias{Accumulator-class}
\title{Example Accumulator class for Emil.}
\description{
An example class for the Emil OOP system.
}
\section{Fields}{
  \describe{
     \item{\code{sum}}{The current amount of the accumlator values}
  }
}
\section{Methods}{
\itemize{
   \item \code{add(x=1)}: Add the value of x to the current accumulated value in sum
}}
\examples{
 x = Accumulator$new()
 x$add(5)
 x$sum
}
```

The methods can be as well documented like this but the dollar should be replaced then by an underline:

```
\name{Accumulator_add}
\alias{Accumulator_add}
\alias{Accumulator$add}
\title{Add the value of x to the current accumulated value in sum.}
\description{This is a method of the Accumulator class which adds the given value to the current object}  
\usage{`Accumulator$add(x=1)`}
\arguments{
    \item{x}{the value to be added to the current internal accumulator value sum, default: 1}
}
\value{object itself invisibly}    
\examples{
 y = Accumulator()
 y$add(5)
 y$sum
 y$add(3)$add(4)
 y$sum
}
```

To see the appropiate manual pages look at the files
`man/Accumulator-class.Rd` and `man/Accumulator_add.Rd`.


